Applying Spatio-temporal Scan Statistics and Spatial Autocorrelation Statistics to identify Covid-19 clusters in the world.

### 2021-01-12 Notebooks/paper_covid.rmd
Update paper_covid.rmd

 At the end of 2019, the situation of infections by Covid-19 *(Sars-Cov-2)* was already being considered an epidemic in the city of Wuhan in China $\!\!^1$, although it was difficult at first to identify the virus' origin, it's known that the Severe Acute Respiratory Syndrome *(SARS)* and the Middle East Respiratory Syndrome *(MERS)* usually need intermediate hosts $\!\!^2$, there is a possibility that bats and pangolins were the first carriers. After several public organizations called for caution relating to Covid-19, on March 12 of 2020 the pandemic state was recognized by the World Health Organization *(WHO)* $\!\!^3$.

@@

  The prospective scan statistic proposed by Kulldorff $\!\!^5$ was chosen to identify emerging spatio-temporal *clusters* in the cumulative number of cases for each month, to apply the technique it was considered that the data follows a Poisson distribution. The purely spatial scan statistic $\!\!^1\!^1$ creates a wide variety of circular windows in different locations throughout the map, each one with a specific group of neighboring areas, all of these circular windows are also flexible in their two dimensions, size and locality, and may or not contain a *cluster* of events. A scan statistic is defined as the likelihood ratio over all possible circles and when maximized identifies the one which constitutes the most likely *cluster*, the *p*-value is obtained through the *Monte-Carlo* hypothesis test.

@@

## Discussion
(here)

@@

 (Here)

@@

[7] - BRAGA, Altemir da Silva. et al. Estudo de Dependência Espacial utilizando Análise de Dados de Área Aplicada na Mesorregião Metropolitana de Belo Horizonte por meio do Indicador Econômico PIB. 19ª SINAPE.  [S.l.]. 2010,  Disponível em: <http://www2.ime.unicamp.br/sinape/sites/default/files/Resumo%20expandido%20SINAPE.pdf>. Acesso em: 24/11/2020.

@@

>>>>

 At the end of 2019, the situation of infections by Covid-19 *(Sars-Cov-2)* was already being considered an epidemic in the city of Wuhan in China $\!\!^1$, although it was difficult at first to identify the virus' origin, it's known that the Severe Acute Respiratory Syndrome *(SARS)* and the Middle East Respiratory Syndrome *(MERS)* usually need intermediate hosts $\!\!^2$, there is a possibility that bats and pangolins were the first carriers. After several public organizations called for caution relating to Covid-19, on March 11 of 2020 the pandemic state was recognized by the World Health Organization *(WHO)* $\!\!^3$.

@@

  The prospective scan statistic proposed by Kulldorff $\!\!^5$ was chosen to identify emerging spatio-temporal *clusters* in the cumulative number of cases for each month, to apply the technique it was considered that the data follows a Poisson distribution. The purely spatial scan statistic creates a wide variety of circular windows in different locations throughout the map, each one with a specific group of neighboring areas, all of these circular windows are also flexible in their two dimensions, size and locality, and may or not contain a *cluster* of events. A scan statistic is defined as the likelihood ratio over all possible circles and when maximized identifies the one which constitutes the most likely *cluster*, the *p*-value is obtained through the *Monte-Carlo* hypothesis test.

@@

## Discussion

  In April global cases of Covid-19 reached 1 million, in the same month the world also reached 3 million cases $\!\!^1\!^0$, with an exponential growth in the number of infections, the Global Moran's I was also very high for the number of deaths per inhabitant, while local spatial association was strong only in Europe, the boxmap **(Picture 3)** shows that all areas identified with local dependence for the number of deaths per inhabitant were either HH (High-High) clusters in April or HL (High-Low) clusters after April, which means that generally locations with a high number of deaths were close to other areas that also had a high number of deaths.

@@

@@

[7] - BRAGA, Altemir da Silva.*et al*. Estudo de Dependência Espacial utilizando Análise de Dados de Área Aplicada na Mesorregião Metropolitana de Belo Horizonte por meio do Indicador Econômico PIB. 19ª SINAPE.  [S.l.]. 2010,  Disponível em: <http://www2.ime.unicamp.br/sinape/sites/default/files/Resumo%20expandido%20SINAPE.pdf>. Acesso em: 24/11/2020.

@@

[10] - Brodeur, *et al*. (2020): A Literature Review of the Economics of COVID-19, GLO Discussion Paper, No. 601, GlobalLabor Organization (GLO), Essen. Available at: http://hdl.handle.net/10419/222316
